`respR` - An R package for the analysis of respirometry data

### 2017-10-18 manuscript/manuscript-HTML.Rmd
manuscript update

---
title: "`respR` - An R package for the analysis of respirometry data"
author: "Januar Harianto, Nicholas Carey, Maria Byrne"
output: 
  html_document:
    toc: false
    number_sections: true
    fig_caption: true
    pandoc_args: ["-F","pandoc-crossref"]
    theme: "united"
---

@@

There are at least four broad methodological approaches in respirometry: closed, intermittent flow, flowthrough and open. In closed respirometry, $O_2$ decrease is measured within a hermetically sealed chamber of known volume, sometimes set within a closed loop to allow circulation or mixing of the environment within the chamber. The oxygen recordings may be continuous through use of an oxygen probe, periodic through withdrawing water or gas samples at set intervals, or a two-point measurement consisting of the initial and final concentrations. Metabolic rates are estimated from respirometry by assuming a linear relationship between variables, and estimates of metabolic rate are straightforward in constant volume respirometry using the equation: $$RO_2 = \Delta O_2(V)$$ where $\Delta O_2$ is the slope of the regression that relates O~2~ concentration to time, or in the case of a two-point measurement, the difference in $O_2$ concentration divided by time, and $V$ is the volume of the container (Lighton, 2008).

@@

A final method is open respirometry, in which an open tank or semi-enclosed area is used, but the input or mixing rate of oxygen from the surroundings is known or found to be negligible relative to oxygen consumption of the specimens (Leclercq et al., 1999). It is seldom used, but for some applications it is a sufficient and practical methodology (e.g. Gamble et al., 2014). The common equation used for open respirometry is: $$RO_2 = \Delta O_2(V) + \phi_d$$ where $\Delta O_2$ is the slope of the regression that relates $O_2$ concentration to time, $V$ is the volume of the container and $\phi_d$ is the air-sea oxygen flux as determined by Fick's Law (Leclercq et al., 1999).

@@

Here we describe the R package `respR`, a set of functions desgined to provide a structured workflow for the analysis of respirometry-related data (Fig. 1). The package contains utilities to: (1) analyse closed, intermittent, flow-through and open respirometry data, (2) scale $\Delta O_2$ to volume and/or mass, and (3) automatically and rapidly detect MMR, SMR, $P_{c}$ or best-fit results in large datasets with the help of traditional rolling regression and kernel density estimation techniques. Other smaller, but useful, functions are also avaliable, and are described in more detail in our online html vignette (https://januarharianto.github.io/respR/). While the package has a strong focus on marine and freshwater respirometry, the principals also apply to aerial respirometry. Some oxygen concentration conversions require a salinity value, but in freshwater respirometry this should be set to zero. We demonstrate the utility of the package by analysing three datasets as case studies and proof of concept. These data were collected using closed, intermittent and flow-through respirometry.

@@

To explore `respR` and its main functions we have provided example data that can be called directly after the package is loaded. The first data object, `urchin2013`, contains measurements of oxygen consumption in 16 individual sea urchins (*Heliocidaris erythrogramma*) and 2 "blank" measurements of background respiration (Harianto, *unpublished*). The `intermittent` data object contains three repeated measures of oxygen consumption in *H. erythrogramma*, collected in a single session with intermittent respirometry (Carey et al. 2016, *JEB*). The `flowthrough` data object contains measurements of oxygen consumption, via flow-through respirometry, in a species of chiton, *Mopalia lignosa* (Carey, *unpublished*). Detailed information about all example data, including their source and methods, can be obtained with the `?` command in the R console (e.g. `?urchin2013`).

@@

Data should be formatted correctly before use in `respR`. The function `check.input()` performs error checks on a data frame of any size, and can extract a two-column data frame for data exploration and subsequent analyses. Data must be numeric, because `respR` has limited support for date/time data (e.g. `POSIXct` and `POSIXlt` classes) and performing calculations on different time formats have many caveats that are beyond the scope of the package. Time data should be sequential, without duplicates, as such errors strongly indicate underlying sampling issues. Ideally, time data should also be evenly-spaced, that is, the time period between samples are equal. This has important ramifications on data extraction techniques later on, as substantial errors may occur in the event that calculations are based on the number of rows instead of time. A plot of the data is also shown for visual inspection (e.g. Fig. 1).

It should be noted that invoking `check.input` is optional -- the main functions in our package will readily accept any data frame as long as data are numeric. Running `check.input` is an exploratory step that flags potential issues about the data before it is further analysed.

## Estimating rates of change in $O_2$ concentration ($RO_2$)

The functions `calc.bg.rate()`, `calc.rate()`, `auto.rate()`, `calc.rate.ft()` and `pcrit()` can process raw respirometry data and are called depending on the experimental setup and type of analysis required.

### Measuring background changes in $RO_2$

Background $RO_2$ is measured to account for the influence of bacterial respiration in the chamber, and once measured, is often applied as a correction to the main experiment (Rodgers et al., 2016). Since background rates typically account for a small percentage of experimental rates, multiple “blank” experiments are routinely conducted alongside main experiments, and the rates are averaged across several datasets to obtain a more accurate estimate of the correction. The function `calc.bg.rate()` uses simple linear regression to simultaneously process multiple background rate measurements, as long as they share the time data, and produces an output object that can be used in later functions that accept background rate arguments.

### Manual and automatic selection of data for calculations

A typical procedure in the analysis respirometry data is to select one or more linear sections of the data for calculations of $RO_2$ (e.g. Gordon et al., 1989, Dorey et al., 2014, Chu & Gal, 2017). The function `calc.rate()` can manually extract one or more segments from a single data frame to process them simultaneously. Data segments can be selected by (1) time period, (2) row numbers, (3) $O_2$ decrease or (4) proportion, which should accomodate most, if not all, data selection requirements and allow for consistent reporting of methods and results (e.g. estimating $RO_2$ over a regular 10-minute time window across multiple specimens). Rates are determined using linear regression, however a simple delta approach is also included alongside all regression results, similar to a two-point analysis which is simply the difference in $O_2$ concentration divided by time.

In addition, we can take advantage of automated methods in `auto.rate()`, which implements rolling regression and kernel density estimation algorithms to detect patterns in the data. From the output we can automatically obtain the maximum or minimum rate of a specified sample size, extract regressions at non-overlapping intervals (Fig. 2a), or detect and rank linear sections of the data by sample size. Unlike manual methods, where data is, likely, arbitrary selected, `auto.rate()` adheres to well-defined rules to obtain the rate of interest, so its methods are results are transparent and fully reproducible. When called, the function takes the data frame of length $m$, and performs a rolling regression of sample size $n$ iteratively across the length of the data (Fig. 2b). Thus, a total of $(m−n)+1$ number of overlapping regressions are fitted, and we obtain a new dataset of regressions as a function of time (rolling regression, Fig. 3), which is then ranked by size to obtain maximum and minimum values.

To estimate the most linear rate, further computations are performed after rolling regressions. Logically, linear sections of the data should reflect a consistently stable (i.e. flat) value in the rolling regression data. The function `auto.rate()` takes advantage of this relationship to identify consistent rates by performing a kernel density estimate of the data, implementing the methods of Sheather & Jones (1991) -- available in the `stats` package in R -- to derive the bandwidth used to construct the "peaks" in the density that correspond to stable rate values. The bandwidth is then used to select the range of overlapping regressions around the peak, and the single section of the data is resampled once using linear regression.

## Flowthrough respirometry

Calculating oxygen consumption using flowthrough respirometry requires different inputs of incurrent and excurrent oxygen concentration and a measure of the rate of flow through the respirometer. The function `calc.rate.ft`

@@

## Converting units of $RO_2$ and $MO_2$

- talk about calc.mo2 and conversions to volume and weight specific $MO_2$.

>>>>

---
title: "`respR` - An R package for the analysis of respirometry data"
author: "Januar Harianto, Nicholas Carey, Maria Byrne"
output: 
  html_document:
    toc: false
    number_sections: true
    fig_caption: true
    theme: "default"
---

@@

There are at least four broad methodological approaches in respirometry: closed, intermittent flow, flowthrough and open. In closed respirometry, $O_2$ decrease is measured within a hermetically sealed chamber of known volume, sometimes set within a closed loop to allow circulation or mixing of the environment within the chamber. The oxygen recordings may be continuous through use of an oxygen probe, periodic through withdrawing water or gas samples at set intervals, or a two-point measurement consisting of the initial and final concentrations. Metabolic rates are estimated from respirometry by assuming a linear relationship between variables, and estimates of metabolic rate are straightforward in constant volume respirometry using the equation: $$VO_2 = \Delta O_2V$$ where $\Delta O_2V$ is the slope of the regression that relates to the rate of change in $O_2$ concentration over time, or in the case of a two-point measurement, the difference in $O_2$ concentration divided by time elapsed, and $V$ is the volume of the container (Lighton, 2008). Note, we use here and in the package the term $RO_2$ to denote rate of $O_2$ depletion or uptake within an experiment. We use this term used to distinguish it from other common terms, for example $VO_2$ which is can be used to denote both volumetric O2 uptake and whole animal uptake, or $MO_2$ which typically denotes mass-specific uptake.

@@

A final method is open respirometry, in which an open tank or semi-enclosed area is used, but the input or mixing rate of oxygen from the surroundings is known or found to be negligible relative to oxygen consumption of the specimens (Leclercq et al., 1999). It is seldom used, but for some applications it is a sufficient and practical methodology (e.g. Gamble et al., 2014). The common equation used for open respirometry is: $$VO_2 = \Delta O_2V + \phi_d$$ where $\Delta O_2V$ is the slope of the regression that relates $O_2$ concentration to time, $V$ is the volume of the container and $\phi_d$ is the air-sea oxygen flux as determined by Fick's Law (Leclercq et al., 1999).

@@

Here we describe the R package `respR`, a set of functions desgined to provide a structured workflow for the analysis of respirometry-related data (Fig. 1). The package contains utilities to: (1) analyse closed, intermittent, flow-through and open respirometry data, (2) scale $\Delta O_2$ to volume and/or mass, and (3) automatically detect MMR, SMR, $P_{c}$ or most linear results in large datasets with the help of traditional rolling regression and kernel density estimation techniques. Other smaller, but useful, functions are also avaliable, and are described in more detail in our online html vignette (https://januarharianto.github.io/respR/). While the package has a strong focus on marine and freshwater respirometry, the principals also apply to aerial respirometry. Some oxygen concentration conversions require a salinity value, but in freshwater respirometry this should be set to zero. We demonstrate the utility of the package by analysing three datasets as case studies and proof of concept. These data were collected using closed, intermittent and flow-through respirometry.

@@

To explore `respR` and its main functions we have provided example data that can be called directly after the package is loaded. Detailed information about all example data, including their source and methods, can be obtained with the `?` command in the R console (e.g. `?urchin2013`). All analyses described in this paper use the example data provided and are fully reproducible (see supplementary information).

@@

Data should be formatted correctly before use in `respR`. The function `check.input()` performs error checks on a data frame of any size, and can extract a two-column data frame for data exploration and subsequent analyses. Data must be numeric, because `respR` has limited support for date/time data (e.g. `POSIXct` and `POSIXlt` classes) and performing calculations on different time formats have many caveats that are beyond the scope of the package. Time data, being a continuous variable, should be sequential, without duplicates. Ideally, time data should also be evenly-spaced, that is, the time period between samples are equal, as this has important ramifications on data extraction methods later on and substantial errors may occur if data is analysed by rows instead of time. The function summarises error checks in the R console (see supplementary material) and plot of the data is shown for visual inspection (e.g. Fig. 1).

It should be noted that invoking `check.input()` is optional -- the main functions in our package will readily accept any data frame as long as data are numeric. Running `check.input` is an exploratory step that flags potential issues about the data before it is further analysed.

## Estimating $RO_2$

The functions `calc.bg.rate()`, `calc.rate()`, `auto.rate()`, `calc.rate.ft()` and `pcrit()` are used to process raw respirometry data. For delibrate reasons, units of measurements are not specified in any of these functions. Instead, they are specified only during unit conversions in `convert.do()`, or when units are scaled to volume and mass in `calc.mo2()`. Separating the

### Analysing background changes in $RO_2$

Background $RO_2$ is often measured to account for the influence of bacterial respiration in the chamber, and applied as a correction to the main experiment (Rodgers et al., 2016). Since background rates typically account for a small percentage of experimental rates, multiple “blank” experiments are routinely conducted and the rates are averaged across several datasets to obtain a more accurate estimate of the correction (e.g. Carey and Sigwart, 2014; Daoud et al., 2007). The function `calc.bg.rate()` uses simple linear regression to simultaneously process multiple background rate measurements and produces an output object that can be used in later functions that accept background rate arguments. Because the oxygen consumption units are non-volume and non-mass specific at the point of correction, the functions in `respR` can adjust $RO_2$ directly without bias: $$RO_{2(correction)} = RO_2 - RO_{2(bg)}$$

### Manual selection of data

Typically, the analysis respirometry data requires the selection one or more linear sections of the raw data for more accurate calculations of $RO_2$ (e.g. Gordon et al., 1989, Dorey et al., 2014, Chu & Gal, 2017). The function `calc.rate()` can manually extract and process data segments from a `data.frame` object. Data segments can be selected by (1) time period, (2) row numbers, (3) $O_2$ decrease or (4) proportion, which should accomodate most, if not all, data selection requirements and allow for consistent reporting of methods and results (e.g. estimating $RO_2$ over a regular 10-minute time window across multiple specimens). Rates are determined using linear regression, however a simple two-point analysis (equation 1) is also included alongside all regression results.

### Automatic selection of data

The function `auto.rate()` uses rolling regression and kernel density estimate algorithms to detect patterns in respirometry data. It can calculate maximum or minimum rates, extract rates at non-overlapping intervals, and has an exploratory method to detect the most linear sections of the data.

There are clear advantages to using `auto.rate()` to estimate MMR and SMR. Unlike manual methods, where data is often arbitrary selected and inconvenient to replicate, `auto.rate()` adheres to well-defined rules to obtain the rate of interest, so its methods are results are transparent and fully reproducible. When called, the function takes the data frame of length $m$, and performs a rolling regression of sample size $n$ iteratively across the length of the data (Fig. 2b). Thus, a total of $(m−n)+1$ number of overlapping regressions are fitted, and we obtain a new dataset of regressions as a function of time (rolling regression, Fig. 3b), which is then ranked by size to obtain maximum and minimum values.

While using rolling regressions to estimate MMR and SMR

### Analysing flowthrough respirometry data

The function `calc.rate.ft()` is used to analyse flowthrough respirometry data, and accepts inputs of incurrent and excurrent oxygen concentration, and flow rate. It is crucial that all the input units match as the function does not perform unit conversions (but see `calc.mo2()`), and the output gives $RO_2$. Like `calc.rate()`, a segment of the data can be selected before processing and the average, or median $VO_2$ are provided.

## Converting units of $RO_2$ to $VO_2$ and $MO_2$

Once

@@

The function `pcrit()` provides a way to consistently report the parameters associated with estimating $P_c$. $P_c$ is often estimated using the "broken-stick" regression (BSR) approach (Yeager and Ultsch, 1989), although there are some cases where it is estimated without indication of how thie value was determined numerically (e.g. need examples). While in many cases the PCrit and oxyregulatory ability can be clear and obvious (e.g./refs), many studies still class species or individuals as oxyregulators or oxyconformers without recognising that there may be intermediate cases.
